---
title: Can You Fold All Your Socks?
author: Package Build
date: '2024-01-07'
slug: can-you-fold-all-your-socks
categories: []
tags: []
editor_options: 
  chunk_output_type: console
---

In my laundry basket, I have 14 pairs of socks that I need to pair up. To do this, I use a chair that can fit nine socks, at most. I randomly draw one clean sock at a time from the basket. If its matching counterpart is not already on the chair, then I place it in one of the nine spots. But if its counterpart is already on the chair, then I remove it from the chair (making that spot once again unoccupied) and place the folded pair in my drawer.

What is the probability I can fold all 14 pairs without ever running out of room on my chair?

Extra credit: What if I change the number of pairs of socks I own, as well as the number of socks that can fit on my chair?
```{r}
library(tidyverse)

simulate_folding <- function(n_pair_socks = 14L, n_chair = 9L) {
    basket_socks <- rep(seq_len(n_pair_socks), 2L)
    chair_socks <- rep(0L, n_chair)
    
    # this runs until we have empty spots on the chair and 
    # we have enough socks to fill the chair
    while (length(basket_socks) > 0 & length(which(chair_socks == 0)) > 0) {
        # place socks on chair
        nsocks_to_sample <- min(length(which(chair_socks == 0)), length(basket_socks))
        socks <- sample(basket_socks, nsocks_to_sample)
        chair_socks[chair_socks == 0] <- socks
        
        # remove socks in chair from total socks (basket)
        basket_socks <- basket_socks[!basket_socks %in% chair_socks]
        # Find duplicate socks on the chair
        dup <- chair_socks[duplicated(chair_socks)]
        
        # remove socks from the chair if duplicate
        chair_socks[which(chair_socks %in% dup)] <- 0L 
    
    }
    # return number of unfolded fold 
    return(tibble(basket_socks = length(basket_socks), 
                  chair_socks = length(which(chair_socks == 0))
                  )
           )
}

simulations_tbl <- tibble(simulation = seq(1, 1e04)) %>% 
    mutate(data = map(simulation, ~ simulate_folding()))  %>% 
    unnest(data)

simulations_tbl %>% 
    #dplyr::filter(basket_socks == 0) %>% 
    summarise(n_unfolded = mean(basket_socks == 0)*100) 
```

```{r} 
```















```{r}
chair <- c(chair,sample(socks, length(which(chair == 0)), replace = FALSE))
n_pair_socks <- 14 
socks <- rep(seq_len(n_pair_socks), 2) 
chair <- sample(socks,9)
while (length(socks) > 0 & length(which(chair == 0)) > 0) { 
    dup <- chair[duplicated(chair)] 
    chair <-chair[!chair %in% dup]
    socks <- socks[!socks %in% dup]
    chair <- c(chair,sample(socks, 9-length(chair), replace = FALSE)) 
}
```
















